---
title: "Bayesian Algorithm for molecular correction"
author: "Veronica Adhiambo"
date: "2025-06-15"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

This notebook runs a Bayesian algorithm to estimate the probability of malaria recrudescence from genotyping data. It is designed for non-technical users and provides a step-by-step walkthrough.

## What you'll need
- An Excel file with the required data (see below for format).
- No need to edit functions.

## What this will do
- Load all necessary packages and sources all our custom R functions from the `R/` directory.
- Load and clean the data from the source codes
- Calls the required functions to:
  - Define alleles
  - Recode alleles
  - Calculate frequencies
  - Run the Bayesian model
  - Output classification tables
- Run the Bayesian model with parallel MCMC 
- Save and plot results for MCMC convergence diagnosis
- Output classification tables


All functions will be available globally for the rest of the analysis.

```{r setup}
library(rJava)
library(gtools)
library(readxl)
library(coda)
library(future)
library(future.apply)
library(abind)
library(gtools)
library(readxl)
library(kableExtra)


scripts <- list.files("functions_scripts/", full.names = TRUE)
sapply(scripts, source, .GlobalEnv)
options(java.parameters = "-Xmx4096m") # Increase Java heap memory  
```


# User defined input

  Edit the following values to match your dataset and desired settings. The R-hat threshold can be maintained at 1.01 and ESS threshold at 400, but you can adjust the chunk size, maximum iterations, and burn-in fraction as needed.
```{r user_defined_input}

# MCMC Chain & Parallel Processing Setup
n_chains <- 4 

#  Automated MCMC Control Parameters

R_hat_threshold <- 1.01
ESS_threshold <- 400
chunk_size <- 2000
max_iterations <- 10000
burn_in_frac <- 0.25


# Locus-specific parameters
# this is the size of the repeat region for each microsatellite (ie. for dimers use 2, for trimers use 3)
locirepeats = c(2,2,3,3,3,3,3)

# this is the tolerance for each marker, used during match counting
bin_sizes = c(0,0,0,0,0,0,0)

# input file which is an excel file with the following sheets: Late Treatment Failures and Additional. You need to change this to the name of your file
inputfile = "Angola_2021_TES_7NMS.xlsx"
```




# Import and Clean Data
load the raw data from the specified Excel file and prepare it for analysis.
```{r import_data, waning=FALSE, message=FALSE} 

# Step 1: We call the function defined in R/import_data.R and pass it the file path. It returns a list containing our two clean data frames.
imported_data <- import_and_clean_data(filepath = inputfile)
genotypedata_latefailures <- imported_data$late_failures
additional_genotypedata <- imported_data$additional
```




# Execute the Analysis Pipeline

This single code chunk executes the entire analysis from start to finish. It performs three main steps:
1.  **Imports and cleans** the raw data from your Excel file using the `import_and_clean_data()` function.
2.  **Runs the core MCMC analysis** for every site using the powerful `run_all_sites()` engine. This step can take several minutes.
3.  **Unpacks the results** into variables that will be used for summary and plotting in the next sections.

```{r excute_mcmc}

# Configure parallel plan
plan(multisession, workers = n_chains)

# Step 2: Execute MCMC for All Sites, Now we pass these clean data frames to our main analysis engine.
analysis_results <- run_all_sites(
    genotypedata_latefailures = genotypedata_latefailures,
    additional_genotypedata = additional_genotypedata,
    locirepeats = locirepeats,
    n_chains = n_chains,
    R_hat_threshold = R_hat_threshold,
    ESS_threshold = ESS_threshold,
    chunk_size = chunk_size,
    max_iterations = max_iterations,
    burn_in_frac = burn_in_frac
)

# MCMC is complete, we can now stop the parallel workers
plan(sequential)


# Step 3: Unpack Results the rest of the script will use these variables.
all_sites_classification <- analysis_results$classifications
all_sites_parameters     <- analysis_results$parameters
all_sites_alleles0       <- analysis_results$alleles0
all_sites_allelesf       <- analysis_results$allelesf
all_sites_ids            <- analysis_results$ids

# The original data is also passed through for the final summary table
genotypedata_latefailures <- analysis_results$genotypedata
```



# Global Summary and output files for the Bayesian approach
 The sites outputs are being merged into one and the final output file is probability_of_recrudescence_summary with summary of the posterior  probabilities of recrudescence for each sample.
 

  This will produce:
  - A histogram of posterior probabilities
  - A final table comparing match-counting and Bayesian classifications
  - Two CSV output files:
    - `probability_of_recrudescence_summary.csv`
    - `classification_output_table.csv`
  
  Color coding
  - Green: Prob. < 0.5 → Likely new infection
  - Orange: Prob. ≥ 0.5 → Likely recrudescence



```{r Summarize Bayesian Results}

if (length(all_sites_ids) == 0) {
    stop("ERROR: No results were generated from any site. Cannot perform final analysis.")
}
list_of_site_dataframes <- lapply(names(all_sites_ids), function(site) {
    site_ids <- all_sites_ids[[site]]
    site_classification <- all_sites_classification[[site]]
    if (is.null(site_ids) || is.null(site_classification) || length(site_ids) != nrow(site_classification)) {
        warning(sprintf("Data for site '%s' is inconsistent or missing. It will be excluded from the final summary.", site))
        return(NULL)
    }
    data.frame(
        Site = site,
        ID = site_ids,
        Probability = rowMeans(site_classification),
        stringsAsFactors = FALSE
    )
})
clean_list_of_dataframes <- Filter(Negate(is.null), list_of_site_dataframes)
if (length(clean_list_of_dataframes) == 0) {
    stop("ERROR: All sites failed processing. No data available for final analysis.")
}
final_summary_df <- do.call(rbind, clean_list_of_dataframes)
hist(final_summary_df$Probability, breaks = 20, 
     main = "Distribution of posterior probability of recrudescence (Successful Sites)", 
     xlab = "Posterior probability of recrudescence")
write.csv(final_summary_df, "probability_of_recrudescence_summary.csv", row.names = FALSE)

genotypedata_latefailures$base_ID <- gsub(" Day 0| Day Failure", "", genotypedata_latefailures$Sample.ID)
matched_indices <- match(genotypedata_latefailures$base_ID, final_summary_df$ID)
probability_of_recrudescence_correctorder <- final_summary_df$Probability[matched_indices]
is_day_failure_row <- grepl("Day Failure", genotypedata_latefailures$Sample.ID)
probability_of_recrudescence_correctorder[is_day_failure_row] <- NA
probability_of_recrudescence_correctorder_formatted <- sapply(probability_of_recrudescence_correctorder, function(x) {
    if (is.na(x)) return("") else return(format(round(x, 2), nsmall = 2))
})
probability_of_recrudescence_correctorder2 <- probability_of_recrudescence_correctorder
probability_of_recrudescence_correctorder2[is.na(probability_of_recrudescence_correctorder2)] <- -1
```




# Generate Final Classification Table
This final step runs the traditional match-counting method and combines its output with the Bayesian probabilities to create a comprehensive comparison table for users.

 

```{r generate_final_table}
match_counting_results <- perform_match_counting(
  genotypedata_latefailures = genotypedata_latefailures,
  bin_sizes = bin_sizes
)

output_table <- cbind(
  genotypedata_latefailures,
  match_counting_results$match_output_table,
  "Number of Matches" = c(rbind(match_counting_results$number_matches, rep("", match_counting_results$nids))),
  "Number of Loci" = c(rbind(match_counting_results$number_loci, rep("", match_counting_results$nids))),
  "Prob. of Recrudescence" = probability_of_recrudescence_correctorder_formatted
)
output_table[is.na(output_table)] <- ""

nloci <- match_counting_results$nloci
locinames <- match_counting_results$locinames
genotype_column_span <- ncol(genotypedata_latefailures) - 2
header_labels <- c(" " = 2, 
                   "Genotypes" = genotype_column_span, 
                   "Individual Marker Calls" = nloci, 
                   "Match Counting" = 2, 
                   "Bayesian" = 1)

# Create the final, beautifully formatted table
finaltable <- kbl(output_table, 
                  caption = "Table. Final classification of recrudescence versus new infection.", 
                  escape = FALSE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 10, full_width = TRUE) %>%
  add_header_above(header_labels, bold = TRUE) %>%
  column_spec(3, border_left = TRUE) %>%
  column_spec(ncol(genotypedata_latefailures) + 1, border_left = TRUE) %>%
  column_spec(ncol(genotypedata_latefailures) + nloci + 1, border_left = TRUE) %>%
  column_spec(ncol(genotypedata_latefailures) + nloci + 3, border_left = TRUE) %>%
  column_spec(ncol(output_table), 
              bold = TRUE,
              background = ifelse(probability_of_recrudescence_correctorder2 < 0.5 & probability_of_recrudescence_correctorder2 >= 0, 
                                  "#C8E6C9", # Softer Green
                                  ifelse(probability_of_recrudescence_correctorder2 >= 0.5, 
                                         "#FFCCBC", # Softer Orange
                                         "#FFFFFF00"))) %>%
  footnote(general = "Classification is based on the Bayesian posterior probability: < 0.5 is considered a New Infection (NI, green), ≥ 0.5 is considered a Recrudescence (R, orange).",
           general_title = "")

write.csv(output_table, "classification_output_table.csv", row.names = FALSE)
finaltable
```



